---
title: "ValidatingSDMSSupplInfo"
author: "Bob O'Hara"
format: html
execute: 
  cache: true

---


```{r, Setup}
#| echo: true
#| message: false
#| cache: false

library(maxnet)
library(disdat)
library(future.apply)
knitr::opts_chunk$set(dpi = 48)
source("ValidatingSDMsFunctions.R")
load("Results/AllCoefs.RData")
 
Regions <- c("AWT", "CAN", "NSW", "NZ", "SA", "SWI")
RemoveNames <- c("siteid", "spid", "x", "y", "occ", "group")

plan(multisession, workers = 6) ## Run in parallel on local computer, with 6 nodes (cores?)

```



## Supplementary Figures

The intercepts from the two models are plotted in @fig-Intercepts. Overall, any correlations are weak.

```{r}
#| label: fig-Intercepts
#| fig-cap: "Estimated intercepts for MaxEnt models with all features fitted to presence-only and presence absence data."

GetInts <- function(region, dat) {
  res <- lapply(dat[[region]], function(lst) {
    df <- c(PO=lst$alpha, PA=c(lst$coefficients["(Intercept)"]))
    df
  })
  res <- t(list2DF(res))
  colnames(res) <- c("PO", "PA")
  res
}

Intercepts <- sapply(Regions, GetInts, dat=AllCoefs.lqpth)
Intercepts.sp <- sapply(Regions, GetInts, dat=AllCoefs.lqpth.sp)

# unlist(lapply(Intercepts, function(x) cor(x)[1,2]))
# unlist(lapply(Intercepts.sp, function(x) cor(x)[1,2]))

par(mfrow=c(2,3), mar=c(3,2,4,1), oma=c(2,2,0,0))
sapply(names(Intercepts), function(nm, lst) {
    plot(lst[[nm]][,"PO"], lst[[nm]][,"PA"], type="n", xlab="", ylab="", main=nm) 
  # rect(-100, -100, 100, 0.7, col="mistyrose", border=NA)
  # rect(-100, -100, 0.7, 100, col="mistyrose", border=NA)
  # rect(-100, -100, 0.7, 0.7, col="pink", border=NA)
  # rect(-100, -100, 100, 0, col="red", border=NA)
  # rect(-100, -100, 0.5, 100, col="red", border=NA)
  points(lst[[nm]][,"PO"], lst[[nm]][,"PA"])
  box()

}, lst=Intercepts)

mtext("Presence Only Data", 1, outer=TRUE)
mtext("Presence Absence Data", 2, outer=TRUE, line=0.5)


```

```{r}
#| label: fig-GoF
#| echo: false
#| results: hide
#| eval: true
#| fig-cap: "Estimated AUC and TSS for calibration of predictions to presence absence data, from Maxent model with all features. Red: values below 0 (TSS) or 0.5 (AUC). Pink: values below 0.4 (TSS) or 0.7 (AUC)"

ExtractGof(lst.sp=AllCoefs.lqpth.sp$AWT$awt01, model="maxnet")

ExtractGof <- function(lst.sp, model="maxnet") {
  c(beta = as.numeric(lst.sp$coefficients["MxPred"]), AUC = lst.sp$valid["AUC", model],
    TSS = lst.sp$valid["TSS", model])
}

GoFstats <- lapply(AllCoefs.lqpth.sp, function(reg) {
  lst <- lapply(reg, ExtractGof)
  res <- as.data.frame(do.call(rbind, lst))
  res
})

PlotTradGoF <- function(nm, lst) {

  plot(lst[[nm]]$AUC, lst[[nm]]$TSS, xlim=c(0.3,1), ylim=c(-0.1,1), type="n", xlab="", ylab="", main=nm) 
  rect(-100, -100, 100, 0.4, col="mistyrose", border=NA)
  rect(-100, -100, 0.7, 100, col="mistyrose", border=NA)
  rect(-100, -100, 0.7, 0.4, col="pink", border=NA)
  rect(-100, -100, 100, 0, col="red", border=NA)
  rect(-100, -100, 0.5, 100, col="red", border=NA)
  points(lst[[nm]]$AUC, lst[[nm]]$TSS)
  box()
}
# PlotTradGoF(nm="CAN", lst=GoFstats)

par(mfrow=c(2,3), mar=c(3,2,4,1), oma=c(2,2,0,0))
sapply(names(GoFstats), PlotTradGoF, lst=GoFstats)
mtext("AUC", 1, outer=TRUE)
mtext("TSS", 2, outer=TRUE, line=0.5)

# lapply(GoFstats, function(x) mean(x$AUC<0.7))
 sum(unlist(lapply(GoFstats, function(x) sum(x$AUC<0.5))))/sum(unlist(lapply(GoFstats, function(x) length(x$AUC))))
 sum(unlist(lapply(GoFstats, function(x) sum(x$TSS<0))))/sum(unlist(lapply(GoFstats, function(x) length(x$AUC))))
```

The traditional goodness of fit statistics are plotted in @fig-GoF. There are weak to moderate correlations between them. Over all species, `r round(100*sum(unlist(lapply(GoFstats, function(x) sum(x$AUC<0.5))))/sum(unlist(lapply(GoFstats, function(x) length(x$AUC)))))`% of species (i.e. `r sum(unlist(lapply(GoFstats, function(x) sum(x$AUC<0.5))))` species) had an AUC below 0.5.





```{r}
#| label: IntCVTSS
#| message: false
#| echo: false
#| results: hide
#| fig-cap: TSS from internal and external validation

  load("Results/CrossValidSpat.RData")
# Clean up NULLs
CVspat.clean <- lapply(CVspat, function(lst) {
  if(is.list(lst)) {
    res <- as.matrix(list2DF(lapply(lst, function(l) if(length(l)==2) {l} else {c(NaN, NaN)})))
    rownames(res) <- names(lst[[1]])
  } else {
    res <- lst
  }
  res
  })

TSS.CV <- sapply(Regions, function(reg, IntCV, valid, stat="TSS") {
  cbind(int=IntCV[[reg]][stat,], ext=valid[[reg]][,stat])
}, IntCV=CVspat.clean, valid=GoFstats)

MinTSS <- min(list2DF(lapply(TSS.CV, function(l) apply(l, 2, min, na.rm=TRUE))))
MaxTSS <- max(list2DF(lapply(TSS.CV, function(l) apply(l, 2, max, na.rm=TRUE))))

par(mfrow=c(2,3), mar=c(3,2,4,1), oma=c(2,2,0,0))
sapply(names(TSS.CV), PlotAUC, lst=TSS.CV, xlim=c(MinTSS,MaxTSS), ylim=c(MinTSS,MaxTSS),
       AtRho = c(0.43, 0.03))
mtext("Internal TSS", 1, outer=TRUE)
mtext("External TSS", 2, outer=TRUE, line=0.5)

```




```{r}
#| label: Diffs
#| fig-height: 12
#| eval: false
#| echo: false

CountObs <- function(region) {
  PO <- disPo(region)
  tab.PO <- table(PO$spid)
  
  if(region%in%c("NSW", "AWT")) {
    grp <- unique(PO$group)
    Count <- sapply(grp, function(G, reg) {
      po <- disPa(reg, group = G)
      apply(po[,!names(po)%in%c("group", "siteid", "x", "y")], 2, sum)
    }, reg=region, simplify = FALSE)
    tab.PA <- unlist(Count)
    names(tab.PA) <- gsub(".*\\.", "", names(tab.PA))
    
  } else {
    PA <- disPa(region)
    tab.PA <- apply(PA[,!names(PA)%in%c("group", "siteid", "x", "y")], 2, sum)
  }
  Counts <- cbind(PA=tab.PA, PO=tab.PO[names(tab.PA)])
  }

AllCounts.l <- sapply(Regions, CountObs)

AllCounts <- data.frame(
  PA = unlist(lapply(AllCounts.l, function(x) x[,"PA"])),
  PO = unlist(lapply(AllCounts.l, function(x) x[,"PO"]))
)
AllCounts$species <- gsub(".*\\.", "", rownames(AllCounts))


AllCorrs <- merge(Corrs.df, CorrsSumm.df, by="species")
AllCorrs$Diff <- AllCorrs$cor - AllCorrs$mean

AllAll <- merge(AllCounts, AllCorrs, by="species")
```

```{r}
#| label: fig-PlotCorrDiff
#| echo: false
#| eval: false
#| fig-height: 6
#| fig-width: 8
#| fig-cap: "Differences between correlations between predictions of Maxent models fitted to presence-only data and presence-absence data, predictions made on presence-absence locations, from actual data and simulated data. Left: differences, right: Correlations plotted agaionst each other"

par(mfrow=c(1,2))
plot(AllCorrs$Diff, AllCorrs$At.region, yaxt="n", xlim=c(-1,2),
     xlab="Difference", ylab="")
segments(AllCorrs$Diff-AllCorrs$sd, AllCorrs$At.region,
         AllCorrs$Diff+AllCorrs$sd, AllCorrs$At.region)

axis(2, at=1:6, labels = levels(AllCorrs$region.x), las=1)
abline(v=0, lty=3)

AllCorrs$RegionCol <- as.numeric(AllCorrs$region.x)
plot(AllCorrs$mean, AllCorrs$cor, pch=16, 
     xlab="Mean of simulated correlations", ylab="Correlation in data", 
     col=AllCorrs$RegionCol)
segments(AllCorrs$mean-AllCorrs$sd, AllCorrs$cor,
         AllCorrs$mean+AllCorrs$sd, AllCorrs$cor, col=AllCorrs$RegionCol)
legend(-0.18, 0.02, levels(AllCorrs$region.x), lty=1, 
       col=1:nlevels(AllCorrs$region.x), cex=0.8)
abline(0,1)


```






```{r}
#| label: fig-PlotCorrDiffNObs
#| fig-height: 12
#| fig-width: 8
#| echo: false
#| eval: false
#| results: hide
#| fig-cap: "Effects of sample size (i.e. number of presences) on correlations between predictions of Maxent models fitted to presence-only data and presence-absence data, predictions made on presence-absence locations"

par(mfrow=c(6,2), mar=c(2,4.1,4.1,1), oma=c(2,2,1,1))
by(AllAll, list(AllAll$region.x), function(df) {
  plot(df$PA, df$mean, log="x", ylab=df$region.y[1])
  if(AllAll$region.x[1]==levels(AllAll$region.x)[1]) mtext("Presence Absence", 3)
  plot(df$PO, df$mean, log="x", xlab="", ylab="")
  if(AllAll$region.x[1]==levels(AllAll$region.x)[1]) mtext("Presence Only", 3)
})
mtext("Number of presences", 1, outer=TRUE)
mtext("Mean correlation in simulations", 2, outer=TRUE)

```

Correlations with bias correction

```{r}
#| label: fig-Corrslqpt.sp
#| results: hide
#| echo: false

AllPreds.lqpth.sp <- lapply(AllCoefs.lqpth.sp, GetPreds)
AllCorrs.lqpth.sp <- lapply(AllPreds.lqpth.sp, CalcCorrs)


Cor.tmp <- unlist(AllCorrs.lqpth.sp)
Corrs.df.sp <- data.frame(cor = Cor.tmp, 
                       region = factor(gsub("\\..*", "", attr(Cor.tmp, "names")), levels=Regions))

par(mfrow=c(1,1), mar=c(4.1,6,1,1))
PlotCorrs(Corrs.df.sp)

```



### AWT

```{r, MaxEntPAPredsAWT}
#| fig-height: 16
#| fig-width: 12
#| results: hide
#| echo: false
#| dev: png
#| fig-cap: "Predictions for presence from models fitted to presence-absence data and presence-only data, for the AWT data"

par(mfrow=c(ceiling(40/6), 6), mar=c(2,2,1,1), oma=c(2,2,0,0))
CorrAWT <- sapply(names(AllCoefs.lqpth$AWT), function(nm, allC) {
  lst <- allC[[nm]]
  plot(lst$pred[,"PA"], lst$pred[,"valid"], xlab="", ylab="", 
       main=nm)
  cor(lst$pred[,"PA"], lst$pred[,"valid"])
}, allC=AllCoefs.lqpth$AWT)
mtext("Fitted to PA data", 1, outer=TRUE)
mtext("Fitted to PO data", 2, outer=TRUE)

```

### CAN


```{r, MaxEntPAPredsCAN}
#| fig-height: 8
#| fig-width: 12
#| eval: true
#| echo: false
#| results: hide
#| dev: png
#| fig-cap: "Predictions for presence from models fitted to presence-absence data and presence-only data, for the CAN data"

par(mfrow=c(ceiling(20/6), 6), mar=c(2,2,1,1), oma=c(2,2,0,0))
CorrCAN <- sapply(names(AllCoefs.lqpth$CAN), function(nm, allC) {
  lst <- allC[[nm]]
  plot(lst$pred[,"PA"], lst$pred[,"valid"], xlab="", ylab="", 
       main=nm)
  cor(lst$pred[,"PA"], lst$pred[,"valid"])
}, allC=AllCoefs.lqpth$CAN)
mtext("Fitted to PA data", 1, outer=TRUE)
mtext("Fitted to PO data", 2, outer=TRUE)

```

### NSW


```{r, MaxEntPAPredsNSW}
#| fig-height: 20
#| fig-width: 12
#| eval: true
#| echo: false
#| results: hide
#| fig-cap: "Predictions for presence from models fitted to presence-absence data and presence-only data, for the NSW data"

par(mfrow=c(11,5), mar=c(2,2,1,1), oma=c(2,2,0,0))
CorrNSW <- sapply(names(AllCoefs.lqpth$NSW), function(nm, allC) {
  lst <- allC[[nm]]
  plot(lst$pred[,"PA"], lst$pred[,"valid"], xlab="", ylab="", 
       main=nm)
  cor(lst$pred[,"PA"], lst$pred[,"valid"])
}, allC=AllCoefs.lqpth$NSW)
mtext("Fitted to PA data", 1, outer=TRUE)
mtext("Fitted to PO data", 2, outer=TRUE)

```

### NZ


```{r, MaxEntPAPredsNZ}
#| fig-height: 20
#| fig-width: 12
#| eval: true
#| echo: false
#| results: hide
#| dev: png
#| fig-cap: "Predictions for presence from models fitted to presence-absence data and presence-only data, for the NZ data"

par(mfrow=c(ceiling(52/6),6), mar=c(2,2,1,1), oma=c(2,2,0,0))
CorrNZ <- sapply(names(AllCoefs.lqpth$NZ), function(nm, allC) {
  lst <- allC[[nm]]
  plot(lst$pred[,"PA"], lst$pred[,"valid"], xlab="", ylab="", 
       main=nm)
  cor(lst$pred[,"PA"], lst$pred[,"valid"])
}, allC=AllCoefs.lqpth$NZ)
mtext("Fitted to PA data", 1, outer=TRUE)
mtext("Fitted to PO data", 2, outer=TRUE)
```

### SA


```{r, MaxEntPAPredsSA}
#| fig-height: 10
#| fig-width: 12
#| eval: true
#| echo: false
#| results: hide
#| dev: png
#| fig-cap: "Predictions for presence from models fitted to presence-absence data and presence-only data, for the SA data"

par(mfrow=c(ceiling(30/6),6), mar=c(2,2,1,1), oma=c(2,2,0,0))
CorrSA <- sapply(names(AllCoefs.lqpth$SA), function(nm, allC) {
  lst <- allC[[nm]]
  plot(lst$pred[,"PA"], lst$pred[,"valid"], xlab="", ylab="", 
       main=nm)
    cor(lst$pred[,"PA"], lst$pred[,"valid"])
}, allC=AllCoefs.lqpth$SA)
mtext("Fitted to PA data", 1, outer=TRUE)
mtext("Fitted to PO data", 2, outer=TRUE)

```

### SWI


```{r, MaxEntPAPredsSWI}
#| fig-height: 10
#| fig-width: 12
#| eval: true
#| echo: false
#| results: hide
#| dev: png
#| fig-cap: "Predictions for presence from models fitted to presence-absence data and presence-only data, for the SWI data"

par(mfrow=c(ceiling(30/6),6), mar=c(2,2,1,1), oma=c(2,2,0,0))
CorrSWI <- sapply(names(AllCoefs.lqpth$SWI), function(nm, allC) {
  lst <- allC[[nm]]
  plot(lst$pred[,"PA"], lst$pred[,"valid"], xlab="", ylab="", 
       main=nm)
  cor(lst$pred[,"PA"], lst$pred[,"valid"])
}, allC=AllCoefs.lqpth$SWI)
mtext("Fitted to PA data", 1, outer=TRUE)
mtext("Fitted to PO data", 2, outer=TRUE)

```




```{r PLotCors}
#| echo: false
#| fig-cap: "More Correlations because we don't have enough"

Corrs <- data.frame(
  Region = rep(1:6, times=c(length(CorrAWT), length(CorrCAN), length(CorrNSW), 
              length(CorrNZ), length(CorrSA), length(CorrSWI))),
    Cor = c(unlist(CorrAWT), unlist(CorrCAN), unlist(CorrNSW), 
            unlist(CorrNZ), unlist(CorrSA), unlist(CorrSWI))
)

par(mfrow=c(1,1), mar=c(4.1,6,1,1))
plot(Corrs$Cor, jitter(Corrs$Region), yaxt="n", xlim=c(-1,1),
     xlab="Correlation", ylab="")
axis(2, at=1:6, labels = Regions, las=1)
abline(v=0, lty=3)
```


